---
title: "ADI exploration example"
author: "Matt Chennells"
date: "2023-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(ggplot2)
  library(microbenchmark)
  library(hrbrthemes)
  library(ineq)
})

here()

# ==========  list data  ==========

rm(list=ls())

d_dec <- read_csv(here("output", "d_deciles.csv"), 
                  col_types = c(year = "f", pop_bins = "f"))

d_full <- read_csv(here("output", "d_full.csv"),
                   col_types = c(year = "f"))
  

# as.factor(year, pop_bins)

mainColor <- "#2876af"
yearColor <- "#69b3a2"
lcColor <- "#b31307"

# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-


# ==========  Gini  ==========

d_dec_lc <- d_dec |> 
  select(year, pop_bins, lc_p, starts_with("lc_L")) |>
  rename(lc_L_ADI = lc_L) |>
  pivot_longer(cols = c(-year, -pop_bins,-lc_p), names_to = "Type", values_to = "Lc", names_prefix = "lc_L_")

d_dec_lc |> count(Type)

## ----- Plot Lorenz curve

# Lc: ADI by year
f_lc <- ggplot(
  data = d_dec_lc |> filter(Type == "ADI"), 
  aes(y = Lc, x = lc_p, group = year, color = year))  +
  geom_line() +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, col = lcColor, lwd = 1) +
  scale_y_continuous(name = "cumulative normalised ADI cases", breaks = seq(0, 1, 0.1)) +
  scale_x_continuous(name = "cumulative normalised rank of ADI cases", breaks = seq(0, 1, 0.1)) +
  coord_cartesian(ylim = c(0,1), xlim = c(0,1)) +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = mainColor, size = 20),
    axis.title.x = element_text(color = yearColor, size = 20),
    legend.position = c(0.9, 0.25) 
  ) +
  ggtitle("Lorenz curve of ADI cases, over time")
f_lc

# But can have similar Gini coefficient but due to either poor getting better off or rich getting worse off

# Lc: ADI by category, most recent year
most_recent_year <- "2021"
d_dec_lc_mry <- d_dec_lc |> filter(year == most_recent_year)

f_lc_cat <- ggplot(
  data = d_dec_lc_mry |> filter(Type == "ADI"), 
  aes(y = Lc, x = lc_p, group = 1))  +
  geom_line(color = mainColor) +
  geom_point(color = mainColor) +
  geom_line(data = d_dec_lc_mry |> filter(Type != "ADI"), 
            aes(group = Type, color = Type), 
            linetype = 3, size = 2) +
  geom_abline(intercept = 0, slope = 1, col = lcColor, lwd = 1) +
  scale_y_continuous(name = "cumulative normalised ADI cases", breaks = seq(0, 1, 0.1)) +
  scale_x_continuous(name = "cumulative normalised rank of ADI cases", breaks = seq(0, 1, 0.1)) +
  coord_cartesian(ylim = c(0,1), xlim = c(0,1)) +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = mainColor, size = 20),
    axis.title.x = element_text(color = yearColor, size = 20),
    legend.position = c(0.8, 0.2) 
  ) +
  ggtitle(paste0("Lorenz curve of ADI cases in ", most_recent_year, ", by category"))
f_lc_cat

# \\ health inequality is low; gini is driven by claims (and then crime) inequality

# ----- compute Gini coefficient
year_list_gini <- unique(d_dec$year)
year_list_gini

d_dec_gini <- tibble()

for (y in year_list_gini) {
  
  gini_ADI <- ineq::Gini( (d_dec |> filter(year == y))$cases_group[-1] )
  gini_claims <- ineq::Gini( (d_dec |> filter(year == y))$cases_group_claims[-1] )
  gini_crime <- ineq::Gini( (d_dec |> filter(year == y))$cases_group_crime[-1] )
  gini_health <- ineq::Gini( (d_dec |> filter(year == y))$cases_group_health[-1] )
  
  z_gini <- tibble(
    year = y, gini_ADI = gini_ADI, gini_claims = gini_claims, gini_crime = gini_crime, gini_health = gini_health
  )
  
  d_dec_gini <- bind_rows(d_dec_gini, z_gini)
  
}

d_dec_gini_long <- d_dec_gini |>
  pivot_longer(cols = c(-year), names_to = "Type", values_to = "gini", names_prefix = "gini_")

## Gini coefficient over time
f_gini_year <- ggplot(
  data = d_dec_gini_long |> filter(Type == "ADI"), 
  aes(y = gini, x = year, group = 1)) +
  geom_line(size = 2, color = mainColor) +
  coord_cartesian(ylim = c(0,0.6)) +
  scale_y_continuous(name = "Gini coefficient") +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = mainColor, size = 20),
    axis.title.x = element_text(color = yearColor, size = 20)
  ) +
  ggtitle("Gini coefficient of ADI cases, by year")
f_gini_year

# >> add labels: % change each year; overall % change

## Gini coefficient over time, by category too
f_gini_year_cat <- ggplot(
  data = d_dec_gini_long |> filter(Type == "ADI"), 
  aes(y = gini, x = year, group = 1)) +
  geom_line(size = 2, color = mainColor) +
  geom_line(data = d_dec_gini_long |> filter(Type != "ADI"), 
            aes(group = Type, color = Type), 
            linetype = 3, size = 2) +
  coord_cartesian(ylim = c(0,0.6)) +
  scale_y_continuous(name = "Gini coefficient") +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = mainColor, size = 20),
    axis.title.x = element_text(color = yearColor, size = 20), 
    legend.position = c(0.9, 0.8)
  ) +
  ggtitle("Gini coefficient of ADI cases by category, by year")
f_gini_year_cat


# ----- add absolute values to the coefficient graph

d_dec_check <- d_dec |> mutate(cases_check = (cases_group_claims + cases_group_crime + cases_group_health ) - cases_group)
# Tiniest discrepancy due to rounding

d_abs <- d_dec |>
  filter(cases_group != 0) |>
  group_by(year) |>
  summarise(
    ADI = sum(cases_group),
    Claims = sum(cases_group_claims),
    Crime = sum(cases_group_crime),
    Health = sum(cases_group_health)
  ) |>
  mutate(across(-year, ~ .x /1000000))

d_abs_check <- d_abs |> mutate(cases_check = (Claims + Crime + Health ) - ADI)


###
d_abs_pivot <- d_abs |> pivot_longer(cols = -year, names_to = "Type", values_to = "N")
d_abs_ADI <- d_abs |> mutate(ADI = Claims + Crime + Health) |> select(year, ADI)
d_abs_CAT <- d_abs |> select(-ADI) |> pivot_longer(cols = -year, names_to = "Type", values_to = "N")

# ADI total and by category, over time
f_cases_year <- ggplot(
  data = d_abs_ADI, 
  aes(y = ADI, x = year, group = 1)) +
  geom_line(size = 2, color = mainColor) +
  geom_line(data = d_abs_CAT,    # Can use d_abs_pivot to include ADI as well
            aes(y = N, x = year, group = Type, color = Type), 
            linetype = 3, size = 2) +
  scale_y_continuous(name = "cases, in millions") +
  scale_x_discrete(name = "year") +
  coord_cartesian(ylim = c(0, 50)) +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = mainColor, size = 20),
    axis.title.x = element_text(color = yearColor, size = 20),
    legend.position = c(0.1,0.8) 
  ) +
  ggtitle("ADI total cases by year, and by category")
f_cases_year


# Proportions shown in stacked graph
f_cases_stack <- ggplot(
  data = d_abs_ADI,  aes(y = ADI, x = as.numeric(year)), group = 1) +
  geom_line(color = mainColor, size = 2) +
  geom_area(
    data = d_abs_CAT,
    aes(y = N, x = as.numeric(year), fill = Type),
    alpha = 0.4
  ) +
  geom_line(data = d_abs_CAT,  aes(y = N, x = as.numeric(year), group = Type, color = Type), linetype = 3, size = 2) +
  scale_x_continuous(name = "year", breaks = seq(2013, 2021, 1)) +
  scale_y_continuous(name = "cases, in millions") +
  coord_cartesian(ylim = c(0, 50)) +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = mainColor, size = 20),
    axis.title.x = element_text(color = yearColor, size = 20),
    legend.position = c(0.1,0.8) 
  ) +
  ggtitle("ADI total cases by year, and by category (shaded = proportion)")
#geom_line( color = mainColor), size = 2)
f_cases_stack

# https://r-graph-gallery.com/stacked-area-graph.html (see also for plotly and dygaph interactive code)

# ----------- By deciles

d_bins <- d_dec |> filter(cases_group != 0) |> 
  mutate(across(starts_with("cases_"), ~ .x / 1000000))

# ADI total and by category, over time
f_cases_bins <- ggplot(
  data = d_bins, 
  aes(y = cases_group, x = year, group = pop_bins, color = pop_bins)) +
  geom_line(size = 2) +
  # geom_line(data = d_abs_CAT,    # Can use d_abs_pivot to include ADI as well
  #           aes(y = N, x = year, group = Type, color = Type), 
  #           linetype = 3, size = 2)
  scale_y_continuous(name = "cases, in millions") +
  scale_x_discrete(name = "year") +
  labs(color = "ADI decile (1=most deprived)") +
  coord_cartesian(ylim = c(0, 10)) +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = mainColor, size = 20),
    axis.title.x = element_text(color = yearColor, size = 20),
    legend.position = c(0.4,0.7)
  ) +
  ggtitle("ADI total cases by ADI decile, by year")
f_cases_bins

# focus on most vulnerable group

d_bins_facet <- d_bins |> 
  select(starts_with("cases_"), year, pop_bins, -cases_group) |>
  pivot_longer(cols = c(-year, -pop_bins), names_to = "Type", values_to = "N", names_prefix = "cases_") |>
  mutate(Type = str_remove(Type ,"group_"))

# ADI category for most deprived groups
f_cases_bin_facet <- ggplot(
  data = d_bins_facet, 
  aes(y = N, x = year, group = Type, color = Type)) +
  geom_line(size = 2) +
  scale_y_continuous(name = "cases, in millions") +
  scale_x_discrete(name = "year") +
  coord_cartesian(ylim = c(0, 6)) +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = mainColor, size = 20),
    axis.title.x = element_text(color = yearColor, size = 20),
    legend.position = c(0.95,0.15) 
  ) +
  ggtitle("ADI cases by ADI decile (1=most deprived), by category and year") +
  facet_wrap(vars(pop_bins))
f_cases_bin_facet


## What about just general rates over time? Should be similar to graph, just moderated by population changes too

# =-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-

# Looking at individual district progress over time, ideas: 
# > look at position changes, either by rank or pop bin change each year
# > likelihood of moving up or down

## can't rank by LSOA as not the case that each LSOA (indicated by 4-digit code) has data for each year

## Looking at area ADI rate rank change by year
d_prog_AREA <- d_full |>
  select(area_code, area_name, pop, ADI_rate, year) |> 
  arrange(area_name) |>
  
  # AREA RATE: collapse into areas, but with ADI rate per area weighted by different LSOA sizes
  mutate(
    area_name_sub = str_sub(area_name, start = 1, end = -5),
  ) |>
  group_by(area_name_sub, year) |>
  mutate(
    pop_area = sum(pop),
    pop_area_weight = pop / pop_area,
    ADI_rate_weight = ADI_rate * pop_area_weight,
    area_ADI_rate = sum(ADI_rate_weight, na.rm = TRUE)   # Some areas missing information on some categories; not many, but make sure to check
  ) |>
  ungroup() |>
  
  # get area-wide rate
  select(area_name_sub, pop, area_ADI_rate, year) |>
  group_by(area_name_sub, year) |>
  summarise(
    area_ADI_rate = max(area_ADI_rate),   # could slice, but just to be safe
    area_pop_total = sum(pop)
  ) |>
  ungroup() |>
  group_by(year) |>
  mutate(
    pop_total = sum(area_pop_total),
    area_pop_total_weight = area_pop_total / pop_total,
    area_ADI_rate_popweighted = area_ADI_rate * area_pop_total_weight
  ) |>
  ungroup()

# >> issues: Coventry has rank = 1 for 2017 but that's because it's missing information there, so more cleaning in future
# microbenchmark if row_number is quicker than: rank; and arrange by year and row number
# >area_rank = rank(area_ADI_rate),
# >arrange(year, area_ADI_rate) |>
#   group_by(year) |>
#   mutate(
#     area_rank_by_year = row_number(),
#     minus_2 = area_rank - area_rank_by_year
#   ) |>
#   ungroup()


## ---------- Rate by ranks over time
# This is different from total counts for equivalently sized population groups; no guarantee that ranked groups will be the same size

d_prog_AREA |> count(year) 

# Average area rate over time (area rates weighted either equally or by pop weight)
d_prog_AREA_calcs <- d_prog_AREA |>
  group_by(year) |>
  summarise(
    ADI_mean = mean(area_ADI_rate),
    ADI_mean_popweighted = sum(area_ADI_rate_popweighted),
    
    ADI_median = median(area_ADI_rate),
    ADI_sd = sd(area_ADI_rate),
  ) |>
  ungroup()
# Can get weighted average if needed

d_prog_AREA_calcs_long <- d_prog_AREA_calcs |> 
  select(-ADI_sd) |> 
  pivot_longer(cols = -year, names_to = "Measure", values_to = "Rate", names_prefix = "ADI_")

f_area_rate_by_year <- ggplot(
  data = d_prog_AREA_calcs_long,
  aes(y = Rate, x = year, group = Measure, color = Measure)
) +
  geom_line(size = 2) +
  scale_y_continuous(name = "ADI rate (average cases/person)") +
  scale_x_discrete(name = "years") +
  coord_cartesian(ylim = c(0.2, 0.8)) +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = mainColor, size = 20),
    axis.title.x = element_text(color = yearColor, size = 20),
    legend.position = c(0.2,0.8)
  ) +
  ggtitle("ADI rate, by year")
f_area_rate_by_year

# Check rank change over time, e.g. 2013-2021, or 2015-2021 (for future ref: some missing data could make this wonky) 
# This is relative: so we might see rank change, but then ask what's happening in absolute terms (e.g. those getting lower rank have greater increase in rate than the decrease in rate experienced by those gaining rank)

#!!! and can check average rate change by those with highest rank changes

# check <- d_prog_AREA |>
#   mutate(across
#          
#          , ~ str_remove(.x, "area_")))

## calculate area ADI_rate rank, by year
prog_bins <- 10

d_prog_AREA_ranks <- d_prog_AREA |>
  group_by(year) |>
  mutate(
    area_rank_by_year = row_number(area_ADI_rate),
    area_rank_by_year_decile = as.factor(cut_interval(area_rank_by_year, n = prog_bins, labels = FALSE))
  ) |>
  ungroup() |>
  mutate(
    area_rank_by_year_decile_band = ifelse(area_rank_by_year_decile %in% c(1,2,3), 1, 
                                           ifelse(area_rank_by_year_decile %in% c(4,5,6,7), 2, 3))
  )

d_prog_AREA_ranks |> count(area_rank_by_year_decile)
d_prog_AREA_ranks |> count(area_rank_by_year_decile_band)

d_prog_AREA_ranks_sum <- d_prog_AREA_ranks |>
  group_by(year, area_rank_by_year_decile) |>
  summarise(
    decile_rate_mean = mean(area_ADI_rate),
    decile_rate_upper = max(area_ADI_rate),
    decile_rate_lower = min(area_ADI_rate),
  )

d_prog_AREA_band_sum <- d_prog_AREA_ranks |>
  group_by(year, area_rank_by_year_decile_band) |>
  summarise(
    decile_rate_mean = mean(area_ADI_rate),
    decile_rate_upper = max(area_ADI_rate),
    decile_rate_lower = min(area_ADI_rate)
  )

f_area_rate_by_year_ranks <- 
  ggplot(
    data = d_prog_AREA_ranks_sum, 
    aes(x = year, y = decile_rate_mean, group = area_rank_by_year_decile, color = area_rank_by_year_decile )
  ) +
  geom_point(size = 2) +
  geom_line(size = 2) +
  geom_line(
    data = d_prog_AREA_ranks,
    aes(y = area_ADI_rate, x = year, group = area_name_sub),
    color = "grey", alpha = 0.2
  ) +
  coord_cartesian(ylim = c(0, 1.5)) +
  scale_y_continuous(name = "average ADI rate") +
  scale_x_discrete(name = "year") +
  labs(color = "ADI rate decile rank") +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = mainColor, size = 20),
    axis.title.x = element_text(color = yearColor, size = 20),
    legend.position = "right"
  ) +
  guides(color = guide_legend(reverse = TRUE)) +
  ggtitle("ADI rate by decile rank, by year")

# theme(legend.position = c(0.2, 0.8))
f_area_rate_by_year_ranks

## ----------- rate changes by rank

# rate changes by band
year_1 <- "2020"
year_2 <- "2021"

# d_prog_AREA_band_sum_change <- d_prog_AREA_band_sum |>
#   filter(year %in% c(year_1, year_2)) |>
#   pivot_wider(names_from = year, values_from = decile_rate_mean, names_prefix = "rate_") |>
#   mutate(rate_change = rate_2021 - rate_2020)

# rate change by decile
d_prog_AREA_ranks_sum_change <- d_prog_AREA_ranks_sum |>
  rename_with(~ str_remove(.x, "decile_rate_"), cols = starts_with("decile_rate_")) |>
  pivot_wider(names_from = year, values_from = c(mean, upper, lower)) |>
  mutate(
    rate_change = mean_2021 - mean_2020,
    rate_color = ifelse(rate_change >= 0, "#69b3a2", "#b31307")
  )

# graph rate changes by rank
f_area_rate_change <- 
  ggplot(
    data = d_prog_AREA_ranks_sum_change, 
    aes(x = area_rank_by_year_decile, y = rate_change, group = rate_color, color = rate_color)
  ) +
  geom_point(size = 2) +
  geom_segment(aes(xend = area_rank_by_year_decile, y = 0, yend = rate_change)) +
  # geom_errorbar() +
  geom_hline(yintercept = 0, size = 0.25, color = "grey") +
  coord_cartesian(ylim = c(-0.02, 0.08)) +
  scale_y_continuous(name = "change in ADI rate") +
  scale_x_discrete(name = "area ADI rate decile rank") +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = mainColor, size = 20),
    axis.title.x = element_text(color = yearColor, size = 20),
    legend.position = "none"
  ) +
  ggtitle(paste0("change in ADI rate between ", year_1," and ", year_2,", by ADI decile rank"))
f_area_rate_change



# Table of rank change

d_prog_AREA_ranks_lag <- d_prog_AREA_ranks |>
  select(area_name_sub, year, area_ADI_rate, area_rank_by_year, area_rank_by_year_decile) |>
  arrange(area_name_sub, year) |>
  group_by(area_name_sub) |>
  mutate (
    rate_change = round(area_ADI_rate - lag(area_ADI_rate), digits = 5),
    rank_change = area_rank_by_year - lag(area_rank_by_year),
    rank_decile_change = as.numeric(area_rank_by_year_decile) - as.numeric(lag(area_rank_by_year_decile))
  ) |>
  ungroup() |>
  rename(
    rate = area_ADI_rate,
    rank = area_rank_by_year,
    rank_decile = area_rank_by_year_decile,
  )

d_prog_AREA_ranks_lag |> count(rank_change)
d_prog_AREA_ranks_lag |> count(rank_decile, rank_decile_change)

# make data wide
d_prog_AREA_ranks_lag_wide <- d_prog_AREA_ranks_lag |>
  pivot_wider(names_from = c(year), values_from = c(rate, rate_change), names_sep = "_")

#  can just focus on the 40 that got better and worse


d_prog_AREA_ranks_lag |> filter(year == "2021") |> count(rank_change)
d_prog_AREA_ranks_lag |> filter(year == "2021") |> count(rank_decile_change)


# View(d_prog_AREA_ranks_lag |> filter(year == "2021", rank_decile_change %in% c(1,-1)))

area_id_rank_change <- d_prog_AREA_ranks_lag |> filter(year == "2021", rank_decile_change %in% c(1,-1))
  area_id_rank_change

d_prog_AREA_ranks_lag_recent <- d_prog_AREA_ranks_lag |> 
  filter(year %in% c("2020","2021"), area_name_sub %in% area_id_rank_change$area_name_sub) |>
  mutate(
    change_group = ifelse(year == "2021" & rank_decile_change == -1, 1, 
                          ifelse(year == "2021" & rank_decile_change == 1, 2,0)),
    rate_color = ifelse(rank_decile_change == 1, "#69b3a2", "#b31307")
  ) |> 
  group_by(area_name_sub) |>
  mutate(
    change_group_max = as.factor(max(change_group))
  ) |>
  ungroup()

# get upper limits of ADI rate decile bands
d_deciles_limits <- d_prog_AREA_ranks |>
  select(year, area_ADI_rate, area_rank_by_year_decile) |> 
  group_by(year, area_rank_by_year_decile) |>
  summarise(
    rank_mean = mean(area_ADI_rate),
    rank_min = min(area_ADI_rate),
    rank_max = max(area_ADI_rate)
  ) |> 
  ungroup()

f_decile_change <- ggplot(
  data = d_prog_AREA_ranks_lag_recent |> filter(change_group_max == 2),
  aes(x = year, y = rate, group = area_name_sub)
  ) +
  geom_point() +
  geom_line() +
  # geom_point(
  #   data = d_deciles_limits,
  #   aes(y = rank_max, group = NULL),
  #   color = "black", size = 4
  # ) +
  geom_errorbar(
    data = d_deciles_limits |> filter(year == "2020"),
    aes(y = NULL, x = year, color = area_rank_by_year_decile, ymin = rank_min, ymax = rank_max, group = NULL),
    size = 2, width = 0.1, alpha = 0.4, position = position_nudge(x = -0.05)
  ) +
  geom_errorbar(
    data = d_deciles_limits |> filter(year == "2021"),
    aes(y = NULL, x = year, color = area_rank_by_year_decile, ymin = rank_min, ymax = rank_max, group = NULL),
    size = 1, width = 0.1, alpha = 0.4, position = position_nudge(x = +0.05)
  ) +
  coord_cartesian(ylim = c(0.2, 1.5)) +
  scale_y_continuous(name = "ADI rate", breaks = seq(0, 25, 0.25)) +
  scale_x_discrete(name = "year") +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = mainColor, size = 20),
    axis.title.x = element_text(color = yearColor, size = 20),
    legend.position = "none",
    panel.grid.major.x = element_blank()
  ) +
  ggtitle("ADI rate in 2020 and 2021 for 40 areas seeing worsening of decile rank (with ADI decile bands shown)")

f_decile_change

d_select_worst <- d_prog_AREA_ranks_lag_recent |> 
  filter(year == "2021") |>
  arrange(desc(rate_change)) |>
  slice_head(n=5)

# select biggest changes only
f_decile_change_worst <- ggplot(
  data = d_prog_AREA_ranks_lag_recent |> filter(area_name_sub %in% d_select_worst$area_name_sub),
  aes(x = year, y = rate, group = area_name_sub)
  ) +
  geom_point() +
  geom_line() +
  annotate("text", x = 2, y = d_select_worst$rate, 
           label = ( paste0(d_select_worst$area_name_sub," (+",round(d_select_worst$rate_change, digits=3),")")), 
           hjust = -1) +
  # geom_point(
  #   data = d_deciles_limits,
  #   aes(y = rank_max, group = NULL),
  #   color = "black", size = 4
  # ) +
  geom_errorbar(
    data = d_deciles_limits |> filter(year == "2020"),
    aes(y = NULL, x = year, color = area_rank_by_year_decile, ymin = rank_min, ymax = rank_max, group = NULL),
    size = 2, width = 0.1, alpha = 0.4, position = position_nudge(x = -0.05)
  ) +
  geom_errorbar(
    data = d_deciles_limits |> filter(year == "2021"),
    aes(y = NULL, x = year, color = area_rank_by_year_decile, ymin = rank_min, ymax = rank_max, group = NULL),
    size = 1, width = 0.1, alpha = 0.4, position = position_nudge(x = +0.05)
  ) +
  coord_cartesian(ylim = c(0.75, 1.25)) +
  scale_y_continuous(name = "ADI rate", breaks = seq(0, 25, 0.25)) +
  scale_x_discrete(name = "year") +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = mainColor, size = 20),
    axis.title.x = element_text(color = yearColor, size = 20),
    legend.position = "none",
    panel.grid.major.x = element_blank()
  ) +
  ggtitle("ADI rate in 2020 and 2021 for 5 areas seeing worst change (with ADI decile bands shown)")

f_decile_change_worst




# print figures



# legend.justification in theme?


```

Annual Deprivation Index (ADI) is a summation of (1) job claims, (2) crime statistics and (3) mental health cases. Combined with data on population per area, this allows us to calculate a rate (ADI cases/person) for each LSOA or Area. As pointed out by the publishers, there are many advantages to having a cardinal scale; the purpose of this doc is to briefly explore these.

## Analysis
One advantage of having ADI rates / area is that we can compare rates across areas. There are many ways to do this. While the IMD involves a relative measure of rank, it carries no information on distribution of impacts. So the first thing we can do is compare rates (so still a relative measure) but look at rate inequality across LSOAs.

The following is a Lorenz curve of ADI rates. It shows cumulative ADI cases (per LSOA, weighted by LSOA population) by cumulative population (ranked by ADI rate, with highest rates to the right hand side), broken into 10 equal population-sized deciles. The 45 degree line represents perfect equality, such that, for example, 20% of the population have 20% of all ADI cases. A curve lying below this reflects the fact that some deciles have more cases than others. 

We can see that there is large inequality in ADI cases. For example, the 60% least affected members of the population experience between 30-40% of all cases. This looks relatively consistent across years.

```{r , echo = FALSE}
f_lc
```

It's useful to see what the major contributor to ADI inequality is. We can eyeball this by looking at inequality among different ADI components. For example, the figure below shows, using 2021 data, inequality in total ADI cases and by ADI component. The biggest inequality is in job claims (curve furthest away from the equality line), followed by crime. It looks like there is not a great (or as great) an inequality in mental health cases.

```{r , echo = FALSE}
f_lc_cat
```

The first graph gave a general idea of inequality in ADI cases, but didn't specifically calculate inequality or make clear any trend in inequality ove time. The following figure shows the Gini coefficient of ADI cases over time. 

The Gini coefficient is a measure of inequality obtained by subtracting the area under the Lorenz curve from the area under the line of equality. Larger Gini coefficients indicate a smaller area under the Lorenz curve (it shifts further away from the line of equality), and so greater inequality.

The figure below shows the change in the Gini coefficient of ADI cases over the data period 2013-2021. It shows that ADI case inequality actually decreased over time, though has increased in the past year. This is good news!

```{r , echo = FALSE}
f_gini_year
```

We can again break ADI total down into its component parts, as in the figure below. The reduction in inequality looks due to reductions in claims and crime inequality, though the relatively large 2016 increase and then subsequent decrease in health inequality is worth mentioning.

```{r , echo = FALSE}
f_gini_year_cat
```

The problem with the analysis above is that inequality decreases only tell us half the story. While reductions over time are good -- they show that there is a slow decrease in ADI inequality -- it ignores changes in overall levels of ADI cases over time. 

The graph below shows clearly that while inequality may have reduced over time, total cases have risen sharply; all groups have seen significant increases in cases (note: to assume rates increase, we must assume that population is relatively constant; I analyse rates shortly). 

The graph also shows that these increases are driven largely by increases in job claims, while crime cases have remained relatively constant. Mental health cases are much lower contributor to ADI cases overall, but notice that they started at a low base and have almost doubled in the past decade.

```{r , echo = FALSE}
f_cases_year
```

Adding in the proportion of all cases by category in the figure below confirms that job claims are the biggest contributor to increases in total ADI cases over time.

```{r , echo = FALSE}
f_cases_stack
```

The figure below shows total cases broken down by LSOA ADI-rate decile (LSOAs ranked by ADI rate, then broken into equal-sized population groups). The top two (of ten) most deprived deciles have a much larger ADI case load than the other deciles; the most deprived has more than double the number of cases than every other decile barring the next most deprived.

```{r , echo = FALSE}
f_cases_bins
```

Putting this in terms of rates makes the comparison even sharper.

It also looks like rates haven't changed much in the past two years, but as we'll see below, there are large differences between the most and least deprived groups -- and a widening of inequality.

```{r , echo = FALSE}
f_area_rate_by_year_ranks
```

And breaking into the various deciles makes the point visually striking; it's particularly the most deprived that have seen the highest change in especially job claims in the past few years.

```{r , echo = FALSE}
f_cases_bin_facet
```

If we look only at the most recent two years of data, we can see that the increase in rates for the two most deprived groups is far higher than the modest changes (and even improvements) in rates for all other groups. This shows a sharp divergence in ADI impacts and a clear widening of inequality.

```{r , echo = FALSE}
f_area_rate_change
```

We can look at those Areas that moved down a decile in ranking. There are 40 of them, and this worsening takes place across the deciles (i.e., not only in the most deprived deciles). The ADI rate bands are shown for clarity.

```{r worst_40, echo = FALSE}
f_decile_change
```

We can identify specific areas that have worsened much more than others. This is useful, as we can then dive into potential reasons why these areas have seen such a sharp change relative to their peers -- and use this information to address social change in something more like real-time than using e.g. only the IMD every few years.

The figue below shows the 5 Areas with the largest increase in ADI rates.


```{r worst_5, echo = FALSE}
f_decile_change_worst
```
